Historical trends in the contribution of six major causes of death to the black-white life expectancy gap by US state

### 2017-04-07 Manuscript/Manuscript_COD_Trends.Rmd

@@

**Methods and Findings**

**Conclusions:**

@@

###Results
```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92

cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
```

```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

@@

```{r meta_regression_of_cod_contribution, echo = F}
#model option 1: FE analysis

#run 12 stratified models. bs() created the b-spline basis
models <- cod_decomp_results %>% 
  group_by(COD, sex) %>%
  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state,
              weights = 1/(COD_cont_yrs_SE)^2, 
              data = .))

#calculated predicted values for all models
dfPred <- broom::augment(models, fit)
dfPred$year <- 1969:2013

#merge the predictions with original dataset
cod_decomp_results <- merge(cod_decomp_results, 
                            dfPred %>% select(COD, sex, year, state, .fitted, .resid), 
                            by = c("COD", "sex", "state", "year")) %>%
                      rename(fitted.FE = .fitted, resid.FE = .resid)

#calculated the sum of the residuals for each state-sex-cod strata (summed over years)
cod_decomp_results <- cod_decomp_results %>% 
  group_by(state, sex, COD) %>%
  mutate(sum.residual = sum(abs(resid.FE)))
```

```{r meta_RE_analysis, echo = F}
library(lme4)

#run 12 stratified models. bs() created the b-spline basis
RE.models <- cod_decomp_results %>% 
  group_by(COD, sex) %>%
  do(fit = lmer(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state),
              weights = 1/(COD_cont_yrs_SE)^2, 
              data = .))

#calculated predicted values for all models
RE.pred <- broom::augment(RE.models, fit)
RE.pred$year <- 1969:2013

RE.models$RE.estimates <- purrr::map(RE.models$fit, ~ ranef(.x))

sex <- rep(c(rep("Female", 40), rep("Male", 40)), 6)
COD <- c(rep("All other causes", 80), rep("Cancers", 80), rep("Cardiovascular", 80),
         rep("Communicable", 80), rep("Injuries", 80), rep("Non-communicable", 80))
RE.estimates <- c(RE.models$RE.estimates[[1]]$state$`(Intercept)`, RE.models$RE.estimates[[2]]$state$`(Intercept)`, 
                  RE.models$RE.estimates[[3]]$state$`(Intercept)`, RE.models$RE.estimates[[4]]$state$`(Intercept)`, 
                  RE.models$RE.estimates[[5]]$state$`(Intercept)`, RE.models$RE.estimates[[6]]$state$`(Intercept)`, 
                  RE.models$RE.estimates[[7]]$state$`(Intercept)`, RE.models$RE.estimates[[8]]$state$`(Intercept)`, 
                  RE.models$RE.estimates[[9]]$state$`(Intercept)`, RE.models$RE.estimates[[10]]$state$`(Intercept)`,
                  RE.models$RE.estimates[[11]]$state$`(Intercept)`, RE.models$RE.estimates[[12]]$state$`(Intercept)`)
state <- rep(levels(cod_decomp_results$state), 12)
REs <- data.frame(sex, COD, RE.estimates, state)

cod_decomp_results <- merge(cod_decomp_results,
                            REs, by = c("COD", "sex", "state"),
                            all.x = T)

#merge the predictions with original dataset
cod_decomp_results <- merge(cod_decomp_results, 
                             RE.pred %>% select(COD, sex, state, year, 
                                                bs.year..degree...1..df...5.,
                                                .fitted, .resid, .fixed, .mu), 
                             by = c("COD", "sex", "state", "year"))

cod_decomp_results <- cod_decomp_results %>% rename(fitted.RE = .fitted, fixed.from.RE = .fixed)

# test1 <- lmer(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + (1|state), weights = 1/(COD_cont_yrs_SE)^2, data = subset(cod_decomp_results, COD == "Cardiovascular" & sex == "Male"))
# 
# pred.general.trend <- predict(test1, re.form = NA)
# pred.with.REs <- predict(test1)
# 
# plot(y=pred.general.trend[1:45], x=1969:2013)
```

@@

  c <- ggplot(subset.data,
              aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey") +
    geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey") +
    geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey") +
    geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey") +
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
    geom_line(col = "#08519c") + 
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs. the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
    theme_classic(base_size = 10) + 
    theme(axis.text.x = element_blank(),
          strip.background=element_blank(),
          axis.line=element_blank(),
          axis.ticks=element_blank()) +
    ylab("Contribution to the life expectancy gap (years)") +
    xlab("Year (1969-2013)")

@@

**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years. The blue band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contribution trends. These curves are overlaid with the meta-estimated common trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).

@@

```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
ggplotly(LE_trend_plot_males)
```

@@

```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
ggplotly(LE_trend_plot_females)
```

@@

```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cardiovascular.Male.plot[[3]]) 
```

@@

```{r print-plot-2, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cardiovascular.Female.plot[[3]]) 
```

@@

```{r print-plot-3, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cancers.Male.plot[[3]]) 
```

@@

```{r print-plot-4, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cancers.Female.plot[[3]]) 
```

@@

```{r print-plot-5, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`Non-communicable.Male.plot`[[3]]) 
```

@@

```{r print-plot-6, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`Non-communicable.Female.plot`[[3]]) 
```

>>>>

```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
```

@@

The overall difference in life expectancy between blacks and whites in the US has decreased over time. However, states have fared differently in their progress towards eliminating the black-white life expectancy gap. Our objective is to describe the pattern of contributions of each of six major causes of deaths to the sex-specific black-white life expectancy gaps over the last half-century, and identify divergent states.

**Methods and Findings**
```{r summarize_results_for_abstract, echo = F}
cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)

ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)

can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(max = round(max(fixed.from.RE), 1))
can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))

inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
```

Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, sex, race, 19 age groups, and six major causes of death. A Bayesian autoregressive model was used to estimate smoothed mortality rates over time and Arriaga’s method was used to decompose the difference in life expectancy. Random effects meta-analysis was used to generate an overall pattern across the states, and was contrasted with the state-specific patterns to identify divergent states.

Although mortality from cardiovascular disease has decreased over time for both racial groups, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r cvd.female.2013` years). Non-communicable diseases’ contribution to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years). Both men and women exhibited a decreased contribution for injuries (men: `r inj.male.1969` to `r inj.male.2013` years, women: `r inj.female.1969` to `r inj.female.2013` years), with the contribution becoming negative for women. Several states did not follow these general trends.

**Conclusions**

The United States has shown improvement in life expectancy for both blacks and whites since 1969. For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further. In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing injury-related mortality among whites. This work emphasizes the importance of understanding how the direction of changes in mortality by race results in changes to the life expectancy gap.

@@

###Results
```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

@@

@@

  c <- ggplot(subset.data,
              aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
    geom_line(col = "#08519c") + 
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs. the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
    theme_classic(base_size = 10) + 
    theme(axis.text.x = element_blank(),
          strip.background=element_blank(),
          axis.line=element_blank(),
          axis.ticks=element_blank()) +
    ylab("Contribution to the life expectancy gap (years)") +
    xlab("Year (1969-2013)")

@@

**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years. The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends. These lines are overlaid with the meta-estimated common trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).

@@

```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
LE_trend_plot_males +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)

```

@@

```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
LE_trend_plot_females +
  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)

```

@@

```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
Cardiovascular.Male.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)

```

@@

```{r print-plot-2, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cardiovascular.Female.plot[[3]]  +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)

```

@@

```{r print-plot-3, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cancers.Male.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)

```

@@

```{r print-plot-4, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cancers.Female.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)

```

@@

```{r print-plot-5, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

```

@@

```{r print-plot-6, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

```
